---
title: "A comprehensive framework for precision medicine with omics data"
author:
- name: Jordi Martorell-Marugán
  affiliation:
  - GENYO, Centre for Genomics and Oncological Research
  - Andalusian Foundation for Biomedical Research in Eastern Andalusia (FIBAO)
  email: jordi.martorell@genyo.es
- name: Daniel Toro-Domínguez
  affiliation:
  - GENYO, Centre for Genomics and Oncological Research
  - Karolinska Institutet
  email: danieltorodominguez@gmail.com
- name: Iván Ellson
  affiliation:
  - GENYO, Centre for Genomics and Oncological Research
- name: Pedro Carmona-Sáez
  affiliation:
  - Department of Statistics and Operational Research. University of Granada
  - GENYO, Centre for Genomics and Oncological Research
  email: pcarmona@ugr.es
package: pathMED
date: "`r BiocStyle::doc_date()`"
abstract: >
  PathMED is a collection of tools to facilitate precision medicine studies
  with omics data (e.g. transcriptomics). Among its funcionalities, genesets
  scores for individual samples may be calculated with several methods.
  These scores may be used to train machine learning models and to predict 
  clinical features on new data. For this, several machine learning
  methods are evaluated in order to select the best method based on internal 
  validation and to tune the hyperparameters. Performance metrics and a 
  ready-to-use model to predict the outcomes for new patients are returned.
output:
  BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{A comprehensive framework for precision medicine with 
  omics data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
references:
- id: toroDominguez2022
  title: "Scoring personalized molecular portraits identify Systemic Lupus Erythematosus subtypes and predict individualized drug responses, symptomatology and disease progression"
  author:
  - family: Toro-Domínguez
    given: D
  - family: Martorell-Marugán
    given: J
  - family: Martínez-Bueno
    given: M
  - family: López-Domínguez
    given: R
  - family: Carnero-Montoro
    given: E
  - family: Barturen
    given: G
  - family: Goldman
    given: D
  - family: Petri
    given: M
  - family: Carmona-Sáez
    given: P
  - family: Alarcón-Riquelme
    given: ME
  container-title: Briefings in Bioinformatics
  issued:
    year: 2022
  volume: 23
  issue: 5
- id: hanzelman2013
  title: "GSVA: gene set variation analysis for microarray and RNA-Seq data"
  author:
  - family: Hänzelmann
    given: S
  - family: Castelo
    given: R
  - family: Guinney
    given: J
  container-title: BMC Bioinformatics
  issued:
    year: 2013
  volume: 14
  issue: 7
- id: barbie2009
  title: "Systematic RNA interference reveals that oncogenic KRAS-driven cancers require TBK1"
  author:
  - family: Barbie
    given: DA
  - family: Tamayo
    given: P
  - family: Boehm
    given: JS
  - family: et al
  container-title: Nature
  issued:
    year: 2009
  volume: 462
- id: foroutan2018
  title: "Single sample scoring of molecular phenotypes"
  author:
  - family: Foroutan
    given: H
  - family: Bhuva
    given: DD
  - family: Lyu
    given: R
  - family: et al
  container-title: BMC Bioinformatics
  issued:
    year: 2018
  volume: 19
- id: tomfohr2005
  title: "Pathway level analysis of gene expression using singular value decomposition"
  author:
  - family: Tomfohr
    given: J
  - family: Lu
    given: J
  - family: Kepler
    given: TB
  container-title: BMC Bioinformatics
  issued:
    year: 2005
  volume: 6
- id: lee2018
  title: "Inferring Pathway Activity toward Precise Disease Classification"
  author:
  - family: Lee
    given: E
  - family: Chuang
    given: HY
  - family: Kim
    given: JW
  - family: et al
  container-title: PLOS Computational Biology
  issued:
    year: 2018
  volume: 4
  issue: 11
- id: aibar2017
  title: "SCENIC: single-cell regulatory network inference and clustering"
  author:
  - family: Aibar
    given: S
  - family: González-Blas
    given: CB
  - family: Moerman
    given: T
  - family: et al
  container-title: Nature Methods
  issued:
    year: 2017
  volume: 14
- id: badia2022
  title: "decoupleR: ensemble of computational methods to infer biological activities from omics data"
  author:
  - family: Badia-i-Mompel
    given: P
  - family: Santiago
    given: JV
  - family: Braunger
    given: J
  - family: et al
  container-title: Bioinformatics Advances
  issued:
    year: 2022
  volume: 2
  issue: 1
- id: korotkevich2021
  title: "Fast gene set enrichment analysis"
  author:
  - family: Korotkevich
    given: G
  - family: Sukhov
    given: V
  - family: Budin
    given: N
  - family: et al
  container-title: bioRxiv
editor_options: 
  markdown: 
    wrap: 72
---

# Install pathMED

You can install pathMED from the GitHub repository:

```{r, eval=FALSE}
devtools::install_github("jordimartorell/pathMED")
```

Please note that on some operating systems such as Ubuntu there may be
installation errors in the packages "metrica", "factoextra" and
"FactoMineR", as it is necessary to install system dependencies with
administrator permissions. These system dependencies can be identified
the following code (adapted to Ubuntu 20.04) and installed by running
the output in a terminal as administrator:

```{r, eval=FALSE}
pak::pkg_system_requirements(c("metrica", "factoextra", "FactoMineR"), 
                             "ubuntu", "20.04")
```

# Core pipeline

(Add figure)

## (Optional) Decomposing gene sets into coexpressed subsets

Gene sets often contain genes associated with a specific biological
function, but their expression can be heterogeneous, sometimes even
exhibiting opposing directions when the function is altered (e.g., in
disease). The `dissectDB()` function clusters the genes within each set
into subsets based on coordinated expression patterns. This clustering
can be performed using one or more molecular datasets, which may include
control samples or not. The same dataset used in the following sections
can also be used for this purpose. Both preloaded and custom gene sets
can be used (see the next section for details). The following code
splits KEGG pathways into subpathways using the example dataset:

The output is a gene set that can be used with the `getScores()`
function. In the new gene set, subpathways are labeled with the suffix
\_splitx. For example, for the pathway hsa04714, the resulting
subpathways will include this suffix.

```{r, message = FALSE, results='hide', warning=FALSE}
library(pathMED)
data(exampleData)

customKEGG <- dissectDB(list(exampleData), geneSets = "kegg")
```

The output is a gene set that can be used with the getScores function.
In the new gene set, subpathways are labeled with the suffix splitx. For
example, for the pathway hsa04714:

```{r, message = FALSE, warning=FALSE}
# Before splitting
print(head(pathMED:::genesetsData[["kegg"]][["hsa04714"]]))

# After splitting
print(customKEGG[grep("hsa04714", names(customKEGG))])

```

Only the genes available in the input dataset are included in this
object.

## Scoring

The `getScores()` function calculates a score for each gene set and
sample. It takes as input a matrix or data frame with molecular features
in rows and samples in columns, along with a gene set collection. The
function supports 22 different scoring methods, which can be specified
using the `method` parameter. The available methods and their references
are:

-   M-score (@toroDominguez2022)
-   GSVA (@hanzelman2013)
-   ssGSEA (@barbie2009)
-   singscore (@foroutan2018)
-   Plage (@tomfohr2005)
-   Z-score (@lee2018)
-   AUCell (@aibar2017)
-   ssGSEA (@barbie2009)
-   AUCell (@aibar2017)
-   decoupleR methods (MDT, MLM, ORA, UDT, ULM, WMEAN, norm_WMEAN,
    corr_WMEAN, WSUM, norm_WSUM and corr_WSUM) (@badia2022)
-   FGSEA and norm_FGSEA (@korotkevich2021)

Gene sets can be provided as a named list, where each entry contains the
genes associated with a specific gene set, using the `geneSets`
parameter. Alternatively, preloaded databases can be used by specifying
their name in the `geneSets` parameter:

-   Reactome ("reactome")
-   KEGG ("kegg")
-   Transcriptional modules associated with immune functions ("tmod")
-   Gene Ontology BP, MF and CC ("go_bp", "go_mf", "go_cc")
-   Disgenet ("disgenet")
-   Human Phenotype Ontology ("hpo")
-   WikiPathways ("wikipathways")
-   PharmGKB ("pharmgkb")
-   LINCS ("lincs")
-   Comparative Toxicogenomics Database ("ctd")

If running the script on a multi-core processor, you can use the `cores`
parameter to specify the number of workers (most `pathMED` functions
include this parameter). Additionally, specific parameters for the
scoring functions can be customized (see, for example,
`?GSVA::gsvaParam`).

To run this function with the example data:

```{r, message = FALSE, results='hide', warning=FALSE}
library(pathMED)
data(exampleData)
scoresExample <- getScores(exampleData, geneSets="kegg", method="GSVA")
head(scoresExample)
```

The result is a matrix with gene set identifiers as rows and samples as
columns. To annotate the identifiers with their corresponding
descriptions, you can use the `ann2term()` function.

```{r, message = FALSE, results='hide', warning=FALSE}
annotatedPathways <- ann2term(scoresExample)
head(annotatedPathways)
```

## Training ML models

The next step in the core `pathMED` pipeline is to use the scores
calculated in the previous step to train machine learning (ML) models
for predicting categorical or numerical clinical variables. The
`trainModel()` function fits and tests multiple algorithms, selecting
the one with the best performance for predicting the specified variable
(`var2predict` parameter). The input objects are `inputData` (the scores
matrix obtained in the previous step), and `metadata` (a data frame
containing sample metadata).

The function `methodsML()` should be used to prepare a list of ML
algorithms to be trained and tested. One or more algorithms can be
selected using the `algorithms` parameter (see the function help for
details), and the amount of hyperparameters combinations to be tested is
specified with the `tuneLength` parameter.

Nested cross-validation is performed using `Koutter` folds for the outer
cross-validation (CV) and `Kinner` folds for the inner CV. The inner
k-fold CV can be repeated multiple times (`repeatsCV` parameter) for
hyperparameter tuning.

In the example dataset, the metadata include a column called `Response`,
indicating whether patients responded to a treatment. The
`positiveClass` parameter can be specified to calculate performance
metrics based on the chosen positive class (e.g., diseased).

```{r, warning=FALSE, eval=FALSE}
data(exampleMetadata)

modelsList <- methodsML(algorithms=c('rf', 'knn'), outcomeClass="character")

trainedModel <- trainModel(scoresExample,
                            metadata=exampleMetadata,
                            var2predict="Response",
                            positiveClass = "YES",
                            models = modelsList,
                            Koutter = 3,
                            Kinner = 3,
                            repeatsCV = 1)

print(trainedModel)
```

The output object is a list with the following elements:

-   **model**: The selected algorithm trained with the entire dataset.
-   **stats**: Performance metrics for each tested algorithms. These
    metrics are calculated with the testing samples of the outter CV.
-   **bestTune**: Optimized hyperparameters for the selected algorithm.
-   **subsample.preds**: The prediction probability for each class and
    sample.

This object may be used with the `predictExternal()` function to make
predictions on new data with the same features than the training data.

# M-scores

In addition to previously developed molecular scoring methods, pathMED
includes the novel M-score method. M-scores were used in a previous
study that used personalized scoring to predict drug response, symptoms
and disease progression in Systemic Lupus Erythematosus patients based
on gene expression data @toroDominguez2022.

M-scores are calculated by comparing pathway activities between cases
and healthy controls, requiring the availability of control sample data.
To use M-scores with the `getScores()` function, the `labels` parameter
must be used to indicate control samples (using 0 or "Healthy").

Using the M-score method, a reference database can be built from
multiple datasets to enable clinical predictions on new samples, even
when healthy control data is unavailable. To achieve this, the following
steps must be followed.

## Build the reference data object

```{r, message = FALSE, results='hide', warning=FALSE}
data(reference_datasets)
```

We loaded `reference_datasets` into the environment, which includes four
gene expression datasets and four corresponding metadata datasets
containing clinical information. These datasets were extracted from NCBI
GEO and contain both lupus and healthy samples, encoded in the metadata
as "Healthy_sample" and "Disease_sample".

The `buildRefObject()` function can be used to create a `refData` object
with the required structure for the following steps. The function
require:

-   A list of expression datasets (`data`) or a single dataset if only
    one is available.
-   A list of metadata files (one per dataset) or a single metadata
    table containing information for all samples.
-   The name of the grouping column in the metadata (`groupVar`).
-   The control group label (`controlGroup`), typically representing
    healthy samples. If different datasets use different control labels,
    a separate value should be provided for each metadata file.

```{r, message = FALSE, results='hide', warning=FALSE}
refData <- buildRefObject(
  data = list(dataset1, dataset2, dataset3, dataset4),
  metadata = list(metadata1, metadata2, metadata3, metadata4),
  groupVar = "group",
  controlGroup = "Healthy_sample")
```

To create the `refData` object, it is recommended to include as many
datasets as possible to improve robustness. Ensure that all gene
expression datasets are log2 transformed and follow a normal
distribution. Additionally, gene names must be annotated as Gene Symbols
in both reference and test datasets.

## Calculate the reference M-scores

After creating a `refData` object, the next step is to calculate
M-scores for the reference datasets with the `mScores_createReference()`
function.

```{r, eval=TRUE, warning=FALSE, message=FALSE, results='hide'}
refMscores <- mScores_createReference(refData,
                           geneSets="tmod",
                           cores=1)
```

## Filter gene sets

To identify the most relevant gene sets for the studied phenotype, you
can use the `mScores_filterPaths()` function. This function selects
pathways based on two key parameters: `perc_samples`, which defines the
percentage of samples in which a pathway must be statistically
significant, and `min_datasets`, which sets the minimum number of
datasets that must meet this th

These parameters help identify pathways that may be significant in small
patient subgroups—a common scenario in heterogeneous diseases—while also
ensuring that the selected pathways appear deregulated across multiple
studies, reducing the risk of study-specific biases. The significance of
each pathway in each sample is determined using its M-score, as
described in the original study @toroDominguez2022.

```{r, warning=FALSE}
relevantPaths <- mScores_filterPaths(MRef=refMscores,
                       min_datasets=3,
                       perc_samples=10)
```

## Calculate the M-scores

Once the reference data is ready and the relevant gene sets selected,
you can calculate the M-scores for your dataset without healthy controls
with the `mScores_imputeFromReference()` function. The input objects are
a matrix with the cases samples, the output of the **diseasePaths**
function. If no healthy controls are provided, the M-scores are
calculated using the *k* most similar samples of the reference data. *k*
may be specified with the **nk** parameter. The output is a matrix with
the M-score of each gene set (rows) and sample (columns).

```{r, warning=FALSE, eval=FALSE}
mScoresExample <- mScores_imputeFromReference(
    inputData=exampleData,
    geneSets=relevantPaths,
    externalReference=refMscores)

print(mScoresExample[1:5, 1:5])
```

# Session info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

# References
